Type theories as quotient inductive-recursive types

### 2025-02-07 tex/main.tex
Polish S4.2

\documentclass[a4paper,UKenglish,numberwithinsect,cleveref,thm-restate]{lipics-v2021}
%This is a template for producing LIPIcs articles. 
%See lipics-v2021-authors-guidelines.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling autoref support, use "autoref"
%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
%for enabling thm-restate support, use "thm-restate"

@@

\supplement{The formal development is hosted at the GitHub repository: \url{https://github.com/genetsai95/DTT-QIIRT}.}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
%\supplementdetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g. Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional
\acknowledgements{We'd like to thank Hsiang-Shang Ko and Meven Lennon-Bertrand.}%optional

@@

\usepackage[obeyFinal,textsize=footnotesize]{todonotes}
\usepackage{manfnt}
\usepackage{underscore}
\newcommand{\danger}{\marginpar[\hfill\dbend]{\dbend\hfill}}
\newcommand{\LT}[2][]{\todo[inline,author={L-T},caption={},#1]{#2}}
\newcommand{\Gene}[2][]{\todo[inline,author={Gene},caption={},#1]{#2}}
\newcommand{\Fredrik}[2][]{\todo[inline,author={Fred},caption={},#1]{#2}}

@@

\begin{abstract}
  \LT[noinline]{Try to interest Thorsten Altenkirch in the abstract}
  Defining type theory in type theory is a long-standing challenge and a benchmark for the expressiveness of a proof assistant.
  Altenkirch and Kaposi used quotient inductive-inductive types to represent exactly the well-typed terms of type theory, with the price to be paid that terms frequently contain transports along equality constructors.
  As a result, proofs need to painstakingly account for coercions alongside the interesting part.
  This %mess
  has been dubbed `transport hell', but is typically suppressed in informal presentation for clarity, hiding the gap between the intention and the formalisation of type theory in type theory.
  In this paper, we aim to shorten this gap, using quotient inductive-inductive-recursive types and definitions by overlapping patterns, reducing the use of transports.
  As a case study, we investigate the (parallel) substitution calculus and type theory with a universe and $\Pi$-types, and compare ours with quotient inductive-inductive definitions.
\end{abstract}

@@

Metaprogramming and mechanised metatheory in a type-theoretic proof assistant relies on having an internal representation of type theory within type theory itself.
%
In the spirit of type theory, we would like to represent well scoped and well typed terms only --- this also has technical advantages, such as theorem statements not having to deal with impossible cases, or separately having to prove wellformedness lemmas, etc.
%
Doing this effectively seems to require several sophisticated features of the host type theory, most notably advanced inductive definitions.
%
For example, inductive families~\cite{Dybjer1994} can be used to represent the simply typed lambda calculus~\cite{Altenkirch1999}, while some kind of simultaneous definitions such as inductive-recursive~\cite{Dybjer2000} or inductive-inductive~\cite{Nordvall-Forsberg2014} definitions seem to be required for dependent type theory~\cite{Danielsson2006,Chapman2009}.
%
This is not just a matter of the pure strength of the host theory, but also expressivity and convenience: while it is known how to reduce inductive-inductive definitions to inductive families in principle~\cite{Kaposi2020a}, the resulting development would be considerable harder to work with, and most likely would have exhausted the resources of both the proof assistant and its user.

@@

\subsection{Definitions by overlapping patterns} \label{sec:meta:overlapping}
\cite{Cockx2014,Altenkirch2016a}
\LT{We use $f\;\vec{x} \reduce \vec{t}$ to indicate a function clause of a definition by overlapping patters.}
\subsection{Quotient inductive-inductive types}
% \subsection{Quotient inductive(-inductive)-recursive types}

@@

A quotient inductive-inductive type~\cite{Altenkirch2018} (QIIT) further allows the simultaneous definition of a type $A$ and a family of types $B : A \to \Set$, where both $A$ and $B$ are defined inductively --- this is different from an inductive-recursive definition of $A$ and $F : A \to D$, where $F$ is defined recursively; indeed, inductive-inductive definitions only make sense when $D = \Set$.
The metatheory of QIITs has been extensively studied by Altenkirch, Kaposi and Kov\'acs~\cite{Kovacs2018,Kaposi2019,Kovacs2020}.

@@

\LT{How the congruence rule for the identity type can be used in QII(R)T?}
\subsection{Definitions by rewrite rules}
\cite{Cockx2020,Cockx2021}

@@

\section{Type theories as quotient inductive-inductive-recursive types} \label{sec:QIIRTs}
\LT{
\begin{enumerate}
  \item Goal: remove transports in the definition so that less transports during reasoning.
  \item Motivation: make type substitution definitional so that we do not have to apply transport along structural rules for types explicitly.
  \item Develop the quotient inductive-inductive-recursive definition of type theory step by step (which needs to be locally confluent and terminating).
    \begin{enumerate}
      \item parallel substitution: type substitution
      \item the type of elements of $A$: term substitution needs to be split into two definitional substitution and explicit term substitution to maintain local confluence.
      \item $\Pi$-type: because of the substitution lifting $\sigma^+ \defeq (\pi_1\idS, \pi_2\idS)$ is used we need to introduce a definitional lifting to maintain local confluence.
        We prefer categorical combinator and avoid using $\left< t \right>$ in the definition, which would introduce transports in the definition.
      \item Other type formers can be introduced as usual (as long as we use categorical combinators).
    \end{enumerate}
\end{enumerate}
}

@@

Substitution calculus has a type $\Ctx$ of contexts, a type $\Ty\,\Gamma$ of types in context $\Gamma : \Ctx$, a type $\Sub\;\Gamma\;\Delta$ of substitutions from context $\Gamma$ to context $\Delta$, and a type $\Tm\;\Gamma\;A$ of terms in context $\Gamma$ of type $A$. 
This amounts to the following type families, all mutually inductively defined (hence inductive-inductive):
\begin{alignat*}{3}
  \Ctx   & : \Set                   \\
  \Ty    & : \Ctx \to \Set          \\
  \Sub   & : \Ctx \to \Ctx \to \Set \\
  \Tm    & : (\Gamma : \Ctx) \to \Ty\;\Gamma \to \Set
\end{alignat*}
The type $\Ctx$ has two constructors: $\emptyctx$ for the empty context, and $\blank,\blank$ for context extension, typed as follows.
\begin{alignat*}{3}
  \emptyctx & : \Ctx \\
  \blank,\blank & : (\Gamma : \Ctx) \to \Ty\;\Gamma \to \Ctx
\end{alignat*}
Note that a context extension $\Gamma , A$ is only wellformed if $A : \Ty\;\Gamma$, ensuring overall wellformedness of contexts, but requiring contexts and types to be defined simultaneously.

Type substitution takes a substitution $\sigma : \Sub\;\Gamma\;\Delta$ and an inhabitant $A : \Ty\;\Delta$ to form a type under $\Gamma$, i.e., type substitution as a constructor has type
\begin{alignat*}{3}
  [\blank]\blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma
\end{alignat*}
where $\Gamma$ and $\Delta$ are kept implicit.
\Fredrik{Could we change this to $A[\sigma]$ instead?}

@@

\LT[noinline]{How about \Coq and \Lean?}
\Fredrik[noinline]{They also translate to eliminators, so I think we do not need to say anything}

@@

As we are defining a function on a quotient inductive type, we have to prove that $[\blank]_{\Ty}\blank$ is coherent, i.e., that it respects all equality constructors.
\LT[noinline]{Should we mention any general condition for coherence?}
\Fredrik[noinline]{Is there one?}
\begin{proposition}[Coherence]\label{prop:coherence-1}
  $[\sigma]_{\Ty}\,A$ is (propositionally) equal to $[\tau]_{\Ty}\,A$ for every equality constructor $p : \sigma = \tau$.
  \danger
\end{proposition}
\begin{proof}
  As equality constructors $[\idS]_T$, $[;]_T$, and $[]\UU$ for types become definitional, it suffices to show that the following identities
  \begin{align*}
    [\sigma;\idS]_{\Ty} \;A          & = [\sigma]_{\Ty}\;A,
                                     & [\idS;\sigma]_{\Ty} \;A          & = [\sigma]_{\Ty}\;A,
                                     & [(\sigma;\tau);\gamma]_{\Ty} \;A & = [\sigma; (\tau; \gamma)]_{\Ty}A, \\
    [\pi_1(\sigma, t)]_{\Ty}\;A      & = [\sigma]_{\Ty}\;A, \\
    [\sigma; (\tau, t)]_{\Ty} \;A    & = [(\sigma;\tau), [\sigma]\, t]_{\Ty}\;A,
                                     & [\sigma]_{\Ty} \;A               & = [\emptyctx]_{\Ty}\;A,
                                     & [\sigma]_{\Ty} \;A               & = [\pi_1\sigma, \pi_2\sigma]_{\Ty}\;A
  \end{align*}
  hold propositionally.
  The first two identities of the first row hold definitionally by \eqref{eq:def-type-subst-1} and \eqref{eq:def-type-subst-2}, while terms of the both sides of the third reduce to $[\sigma]_{\Ty}\;([\tau]_{\Ty}\;([\gamma]_{\Ty}\;A))$ by \eqref{eq:def-type-subst-2}.

@@

Finally, we show that these simultaneously defined functions are coherent.
\begin{proposition}[Coherence]
  For every $p : \sigma = \tau : \Sub\;\Gamma\;\Delta$, the following identities
  \danger
  \[
    \sigma \uparrow A =^{\lam{\gamma}\Sub\;(\Gamma, [ \gamma ]_{\Ty} A)\;(\Delta, A)}_{\alert{p}} \tau \uparrow A
    \qquad\text{and}\qquad
    [\sigma]_{\Tm}\,t =^{\lam{\gamma}\Tm\;\Gamma\;([\gamma]_{\Ty}A)}_{\alert{p}} [\tau]_{\Tm}\,t,
  \]
  hold for any type $A$ and any term $t$.
\end{proposition}
\begin{proof}
  These two identities follow from \cref{prop:correctness-2,prop:correctness-3} easily with the heterogeneous identity (see \cref{sec:meta}).
\Fredrik{This heterogeneous equality has not been used so far; it would be good to not have to introduce it all, except perhaps locally in the proof here.}
\end{proof}

\subsubsection{Multiple lifting and the coherence of type substitution}
\Fredrik{This section gets rather technical. Keeping the page limit in mind, I wonder if we should only briefly summarise it in the main text?}

@@

\subsection{... and other type formers} \label{subsec:SC+U+Pi+more}
\LT{%
\begin{enumerate}
  \item Discuss the extensional identity type.
  \item Discuss Coquand universes.
  \item Discuss the unit type.
  \item Discuss the Boolean type.
  \item Discuss the type of natural numbers.
\end{enumerate}
}
\Fredrik{I think we should keep this brief. I started doing the natural numbers in Agda, but there were some problems with the rewrite $[\sigma] \mathsf{Nat} = \mathsf{Nat}$ not firing in the type of the eliminator...}

@@

To eliminate $\Ctx$, $\Ty\,\Gamma$, $\Sub\;\Gamma\;\Delta$, and $\Tm\,\Gamma\,A$, we have to specify types
\begin{align*}
  \Ctx^M & : \Ctx \to \Set &
  \Sub^M & : \implicit{\Gamma, \Delta}\,\Ctx^M\;\Gamma \to \Ctx^M\;\Delta \to \Sub\;\Gamma\;\Delta \to \Set \\
  \Ty^M  & : \implicit{\Gamma}\,\Ctx^M\;\Gamma \to \Ty\;\Gamma \to \Set & 
  \Tm^M  & : \implicit{\Gamma, A}\; (\Gamma^M : \Ctx^M\;\Gamma) \to \Ty^M\;\Gamma^M\;A \to \Tm\;\Gamma\;A \to \Set
\end{align*}
called the \emph{motivates} as the target types for the elimination:
\begin{alignat*}{5}
  \elim_{\Ctx} & : (\Gamma : \Ctx)   && \to \Ctx^M\;\Gamma \\
  \elim_{\Ty}  & : (A : \Ty\;\Gamma) && \to \Ty^M\;(\elim_{\Ctx}\,\Gamma)\;A \\
  \elim_{\Sub} & : (\sigma : \Sub\;\Gamma\;\Delta) && \to \Sub^M\;(\elim_{\Ctx}\,\Gamma)\;(\elim_{\Ctx}\;\Delta)\;\sigma \\
  \elim_{\Tm}  & : (t : \Tm\;\Gamma\;A) && \to \Tm^M\;(\elim_{\Ctx}\,\Gamma)\;(\elim_{\Ty}\,A)\;t.
\end{alignat*}
Each constructor $c$ is interpreted by a corresponding functions $c^M$ indexed by $c$, i.e.\ 
\begin{alignat*}{3}
  \emptyctx_{\Ctx}^M     & :                      && \Ctx^M\;\emptyctx \\
  \blank,_{\Ctx}^M\blank & : \implicit{\dots} && (\Gamma^M : \Ctx^M\;\Gamma) \to \Ty^M\;\Gamma^M\;A \to \Ctx^M\;(\Gamma, A) \\
  \UU^M & : \implicit{\dots} && \Ty^M\;\Gamma^M\;\UU \\
  \emptysub_{\Sub}^M     & : \implicit{\dots} && \Sub^M\;\Gamma^M\;\emptyctx_{\Ctx}^M\;\cdot \\
  \blank,_{\Sub}^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Gamma^M\;([\sigma^M]^M_{\Ty}\,A^M)\;t \to \Sub^M\;\Gamma^M\;(\Delta^M ,^M_{\Ctx} A^M)\;(\sigma, t) \\
  \idS^M          & : \implicit{\dots} && \Sub^M\;\Gamma^M\;\Gamma^M\;\idS \\
  [\blank]^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]\,t) \\
                   && \vdots
\end{alignat*}
and simultaneously defined functions are also given by corresponding functions 
\begin{alignat*}{3}
  [\blank]_{\Ty}^M\blank & : \implicit{\dots} && \Sub^M\;\Gamma^M\;\Delta^M\;\sigma \to \Ty^M\;\Delta^M\;A \to \Ty^M\;\Gamma^M\;([\sigma]_{\Ty}\,A) \\
  [\blank]_{\Tm}^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]_{\Tm}\,t) \\
                         && \vdots
\end{alignat*}
subject to the corresponding equations \emph{and function clauses}, e.g., 
\begin{alignat*}{5}
  [\idS]_{\Ty}^M & : \implicit{\dots} && [ \idS^M ]^M_{\Ty} \;A^M && =^{\Ty^M\,\Gamma^M}\;&& A^M \\
  [;]_{\Ty}^M & : \implicit{\dots} && [ \sigma^M ;^M \tau^M ]^M_{\Ty} \;A^M && =^{\Ty^M\,\Gamma^M}\; && [\sigma^M]^M_{\Ty}\,([\tau^M]^M_{\Ty}\,A^M) \\
  [\idS]_{\Tm}^M & : \implicit{\dots} && [\idS^M]^M_{\Tm}\;t^M && =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{[\idS^M]}} && t^M \\
                 && \vdots
\end{alignat*}
called \emph{methods}.
For brevity, we omit the rest of methods and refer to the Agda implementation for them.
\LT[noinline]{Add a link.}
We also have omitted the list of implicitly quantified variables: each variable $A$ implicitly quantified in a constructor $c$ has accordingly an $A$-indexed variable $A^M$ in $c^M$.

@@

Given an eliminator, the following clauses are derived to carry out the actual computation
\begin{alignat*}{5}
  \elim_{\Ctx}\,\emptyctx   & \reduce \emptyctx_{\Ctx}                                    && : \Ctx^M\;\emptyctx  \\
  \elim_{\Ctx}\,(\Gamma, A) & \reduce (\elim_{\Ctx}\,\Gamma),_{\Ctx}^{M} (\elim_{\Ty}\,A) && : \Ctx^M\;(\Gamma, A) \\
  \elim_{\Ty}\,\UU          & \reduce \UU^M && :\Ty^M\;\Gamma^M\;\UU \\
  \elim_{\Sub}\,\emptysub   & \reduce \emptysub^M_{\Sub} &&: \Sub^M\;\Gamma^M\;\emptyctx^M_{\Ctx}\;\emptyctx \\
                            & \vdots
\end{alignat*}
interpreting each constructor using the corresponding method inductively.

>>>>

\documentclass[a4paper,UKenglish,numberwithinsect,cleveref,thm-restate,final]{lipics-v2021}
%This is a template for producing LIPIcs articles. 
%See lipics-v2021-authors-guidelines.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling autoref support, use "autoref"
%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
%for enabling thm-restate support, use "thm-restate"

@@

\supplement{The formal development is hosted at the \textsc{GitHub} repository: \url{https://github.com/genetsai95/DTT-QIIRT}.\LT[noinline]{Transfer it to my account?}}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
%\supplementdetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g. Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional
\acknowledgements{We'd like to thank Hsiang-Shang Ko and Meven Lennon-Bertrand.}%optional

@@

\usepackage[obeyFinal,textsize=footnotesize]{todonotes}
\usepackage{manfnt}
\newcommand{\danger}{\marginpar[\hfill\dbend]{\dbend\hfill}}
\newcommand{\LT}[2][]{\todo[inline,author={L-T},caption={},color={pink},#1]{#2}}
\newcommand{\Gene}[2][]{\todo[inline,author={Gene},caption={},#1]{#2}}
\newcommand{\Fredrik}[2][]{\todo[inline,author={Fred},caption={},#1]{#2}}

@@

\begin{abstract}
  Defining type theory in type theory is a long-standing challenge and a benchmark for the strength and expressivity of type theory and a proof assistant.
  Altenkirch and Kaposi used quotient inductive-inductive types to represent exactly the well-typed terms of type theory, with the price to be paid that terms frequently contain transports along equality constructors.
  As a result, proofs need to painstakingly account for coercions alongside the interesting part.
  This %mess
  has been dubbed `transport hell', but is typically suppressed in informal presentation for clarity, hiding the gap between the intention and the formalisation of type theory in type theory.
  In this paper, we aim to shorten this gap, using quotient inductive-inductive-recursive types and definitions by overlapping patterns, reducing the use of transports.
  As a case study, we investigate substitution calculus and type theory with a universe and $\Pi$-types, and compare ours with quotient inductive-inductive definitions.
\end{abstract}

@@

Metaprogramming and mechanised metatheory in a type-theoretic proof assistant relies on having an internal representation of type theory within type theory itself.
%
In the spirit of type theory, we would like to represent well scoped and well typed terms only --- this also has technical advantages, such as theorem statements not having to deal with impossible cases, or separately having to prove wellformedness lemmas, etc.
%
Doing this effectively seems to require several sophisticated features of the host type theory, most notably advanced inductive definitions.
%
For example, inductive families~\cite{Dybjer1994} can be used to represent the simply typed lambda calculus~\cite{Altenkirch1999}, while some kind of simultaneous definitions such as inductive-recursive~\cite{Dybjer2000} or inductive-inductive~\cite{Nordvall-Forsberg2014} definitions seem to be required for dependent type theory~\cite{Danielsson2006,Chapman2009}.
%
This is not just a matter of the pure strength of the host theory, but also expressivity and convenience: while it is known how to reduce inductive-inductive definitions to inductive families in principle~\cite{Kaposi2020a}, the resulting development would be considerably harder to work with, most likely exhausting the resources of both the proof assistant and its user.

@@

\subsection{Quotient inductive-inductive types}
% \subsection{Quotient inductive(-inductive)-recursive types}

@@

A quotient inductive-inductive type~\cite{Altenkirch2018} (QIIT) further allows the simultaneous definition of a type $A$ and a family of types $B : A \to \Set$, where both $A$ and $B$ are defined inductively --- this is different from an inductive-recursive definition of $A$ and $F : A \to D$, where $F$ is defined recursively; indeed, inductive-inductive definitions only make sense when $D = \Set$.
The metatheory of QIITs has been extensively studied by Altenkirch, Kaposi and Kov\'acs~\cite{Kaposi2018,Kaposi2019,Kovacs2020}.

@@

\LT{How the congruence rule for the identity type can be used in QII(R)T?}

\subsection{Definitions by overlapping patterns} \label{sec:meta:overlapping}
\cite{Cockx2014,Altenkirch2016a}
\LT{We use $f\;\vec{x} \reduce \vec{t}$ to indicate a function clause of a definition by overlapping patters.}
\subsection{Definitions by rewrite rules}
\cite{Cockx2020,Cockx2021}

@@

\section{Type theories as quotient inductive-inductive-recursive types} \label{sec:QIIRTs}

@@

Substitution calculus has a type $\Ctx$ of contexts, a type $\Ty\,\Gamma$ of types in context $\Gamma : \Ctx$, a type $\Sub\;\Gamma\;\Delta$ of substitutions from context $\Gamma$ to context $\Delta$, and a type $\Tm\;\Gamma\;A$ of terms in context $\Gamma$ of type $A$. 
This amounts to the following type families, all mutually inductively defined (hence inductive-inductive):
\begin{align*}
  \Ctx   & : \Set                   & \Sub   & : \Ctx \to \Ctx \to \Set \\
  \Ty    & : \Ctx \to \Set          & \Tm    & : (\Gamma : \Ctx) \to \Ty\;\Gamma \to \Set
\end{align*}
$\Ctx$ has two constructors: $\emptyctx$ for the empty context, and $\blank,\blank$ for context extension, typed as
\begin{alignat*}{3}
  \emptyctx & : \Ctx \\
  \blank,\blank & : (\Gamma : \Ctx) \to \Ty\;\Gamma \to \Ctx
\end{alignat*}
Note that a context extension $\Gamma , A$ is only wellformed if $A : \Ty\;\Gamma$, ensuring overall wellformedness of contexts, but requiring contexts and types to be defined simultaneously.

Type substitution takes a substitution $\sigma : \Sub\;\Gamma\;\Delta$ and an inhabitant $A : \Ty\;\Delta$ to form a type under $\Gamma$, i.e., type substitution as a constructor has type
\begin{alignat*}{3}
  [\blank]\blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma
\end{alignat*}
where $\Gamma$ and $\Delta$ are kept implicit.
\Fredrik{Could we change this to $A[\sigma]$ instead?}
\LT{I am okay with changing it in the paper: This notation is used in the formalisation to avoid using $\comp$ for substitution composition: }

@@

@@

As we are defining a function on a quotient inductive type, we have to prove that $[\blank]_{\Ty}\blank$ is coherent, i.e., that it respects all equality constructors.
\LT[noinline]{Should we mention any general condition for coherence?}
\Fredrik[noinline]{Is there one?}
\LT[noinline]{I think Cubical Agda has sort of conditions? (Cubical Agda does not support rewriting)}
\begin{proposition}[Coherence]\label{prop:coherence-1}
  $[\sigma]_{\Ty}\,A$ is (propositionally) equal to $[\tau]_{\Ty}\,A$ for every equality constructor $p : \sigma = \tau$.
  \danger
\end{proposition}
\begin{proof}
  As equality constructors $[\idS]_T$, $[;]_T$, and $[]\UU$ for types become definitional, it suffices to show that the following identities
  \begin{align*}
    [\sigma;\idS]_{\Ty} \;A          & = [\sigma]_{\Ty}\;A,
                                     & [\idS;\sigma]_{\Ty} \;A          & = [\sigma]_{\Ty}\;A,
                                     & [(\sigma;\tau);\gamma]_{\Ty} \;A & = [\sigma; (\tau; \gamma)]_{\Ty}A, \\
    [\pi_1(\sigma, t)]_{\Ty}\;A      & = [\sigma]_{\Ty}\;A, \\
    [\sigma; (\tau, t)]_{\Ty} \;A    & = [(\sigma;\tau), [\sigma]\, t]_{\Ty}\;A,
                                     & [\sigma]_{\Ty} \;A               & = [\emptyctx]_{\Ty}\;A,
                                     & [\sigma]_{\Ty} \;A               & = [\pi_1\sigma, \pi_2\sigma]_{\Ty}\;A
  \end{align*}
  hold propositionally.
  The first two identities of the first row hold definitionally by \eqref{eq:def-type-subst-1} and \eqref{eq:def-type-subst-2}, while terms of the both sides of the third reduce to $[\sigma]_{\Ty}\;([\tau]_{\Ty}\;([\gamma]_{\Ty}\;A))$ by \eqref{eq:def-type-subst-2}.

@@

Finally, we show that these simultaneously defined functions are coherent.
\begin{proposition}[Coherence]
  For every $p : \sigma = \tau : \Sub\;\Gamma\;\Delta$, the following identities
  \danger
  \[
    \sigma \uparrow A =^{\lam{\gamma}\Sub\;(\Gamma, [ \gamma ]_{\Ty} A)\;(\Delta, A)}_{\alert{p}} \tau \uparrow A
    \qquad\text{and}\qquad
    [\sigma]_{\Tm}\,t =^{\lam{\gamma}\Tm\;\Gamma\;([\gamma]_{\Ty}A)}_{\alert{p}} [\tau]_{\Tm}\,t,
  \]
  hold for any type $A$ and any term $t$.
\end{proposition}
\begin{proof}
  These two identities follow from \cref{prop:correctness-2,prop:correctness-3} easily with the heterogeneous identity (see \cref{sec:meta}).
\Fredrik{This heterogeneous equality has not been used so far; it would be good to not have to introduce it all, except perhaps locally in the proof here.}
\LT{How about deferring introducing heterogeneous equality until \cref{sec:formal-implementation}?}
\end{proof}

\subsubsection{Multiple lifting and the coherence of type substitution}
\Fredrik{This section gets rather technical. Keeping the page limit in mind, I wonder if we should only briefly summarise it in the main text?}
\LT{Move this to the appendix?}

@@

\subsection{Adding other type formers} \label{subsec:SC+U+Pi+more}
\LT{%
\begin{enumerate}
  \item Discuss the extensional identity type.
  \item Discuss Coquand universes.
  \item Discuss the unit type.
  \item Discuss the Boolean type.
  \item Discuss the type of natural numbers.
\end{enumerate}
}
\Fredrik{I think we should keep this brief. I started doing the natural numbers in Agda, but there were some problems with the rewrite $[\sigma] \mathsf{Nat} = \mathsf{Nat}$ not firing in the type of the eliminator...}

@@

To eliminate $\Ctx$, $\Ty\,\Gamma$, $\Sub\;\Gamma\;\Delta$, and $\Tm\,\Gamma\,A$, we have to specify types
\begin{align*}
  \Ctx^M & : \Ctx \to \Set &
  \Sub^M & : \implicit{\Gamma, \Delta}\,\Ctx^M\;\Gamma \to \Ctx^M\;\Delta \to \Sub\;\Gamma\;\Delta \to \Set \\
  \Ty^M  & : \implicit{\Gamma}\,\Ctx^M\;\Gamma \to \Ty\;\Gamma \to \Set & 
  \Tm^M  & : \implicit{\Gamma, A}\; (\Gamma^M : \Ctx^M\;\Gamma) \to \Ty^M\;\Gamma^M\;A \to \Tm\;\Gamma\;A \to \Set
\end{align*}
called the \emph{motivates} as the target types for the elimination principle:
\begin{alignat*}{5}
  \elim_{\Ctx} & : (\Gamma : \Ctx)   && \to \Ctx^M\;\Gamma &
  \elim_{\Sub} & : (\sigma : \Sub\;\Gamma\;\Delta) && \to \Sub^M\;(\elim_{\Ctx}\,\Gamma)\;(\elim_{\Ctx}\;\Delta)\;\sigma \\
  \elim_{\Ty}  & : (A : \Ty\;\Gamma) && \to \Ty^M\;(\elim_{\Ctx}\,\Gamma)\;A \quad
               & \elim_{\Tm}  & : (t : \Tm\;\Gamma\;A) && \to \Tm^M\;(\elim_{\Ctx}\,\Gamma)\;(\elim_{\Ty}\,A)\;t.
\end{alignat*}
Each constructor $c$ is interpreted by a corresponding functions $c^M$ indexed by $c$, e.g.,\ 
\begin{alignat*}{3}
  \emptyctx_{\Ctx}^M     & :                    && \Ctx^M\;\emptyctx \\
  \blank,_{\Ctx}^M\blank & : \implicit{\dots}\; && (\Gamma^M : \Ctx^M\;\Gamma) \to \Ty^M\;\Gamma^M\;A \to \Ctx^M\;(\Gamma, A) \\
  \UU^M & : \implicit{\dots} && \Ty^M\;\Gamma^M\;\UU \\
  \emptysub_{\Sub}^M     & : \implicit{\dots}   && \Sub^M\;\Gamma^M\;\emptyctx_{\Ctx}^M\;\cdot \\
  \blank,_{\Sub}^M\blank & : \implicit{\dots}   && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Gamma^M\;([\sigma^M]^M_{\Ty}\,A^M)\;t \to \Sub^M\;\Gamma^M\;(\Delta^M ,^M_{\Ctx} A^M)\;(\sigma, t) \\
%  \idS^M          & : \implicit{\dots} && \Sub^M\;\Gamma^M\;\Gamma^M\;\idS \\
  [\blank]^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]\,t)
\end{alignat*}
and simultaneously defined functions are also given by corresponding functions 
\begin{alignat*}{3}
  [\blank]_{\Ty}^M\blank & : \implicit{\dots} \; && \Sub^M\;\Gamma^M\;\Delta^M\;\sigma \to \Ty^M\;\Delta^M\;A \to \Ty^M\;\Gamma^M\;([\sigma]_{\Ty}\,A) \\
  [\blank]_{\Tm}^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]_{\Tm}\,t)
\end{alignat*}
subject to the identities corresponding to equality constructors and function clauses, e.g., 
\begin{alignat*}{5}
  [\idS]_{\Ty}^M & : \implicit{\dots}\; && [ \idS^M ]^M_{\Ty} \;A^M && =^{\Ty^M\,\Gamma^M}\;&& A^M \\
  [;]_{\Ty}^M    & : \implicit{\dots}   && [ \sigma^M ;^M \tau^M ]^M_{\Ty} \;A^M && =^{\Ty^M\,\Gamma^M}\; && [\sigma^M]^M_{\Ty}\,([\tau^M]^M_{\Ty}\,A^M) \\
  [\idS]_{\Tm}^M & : \implicit{\dots}   && [\idS^M]^M_{\Tm}\;t^M && =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{[\idS^M]}} && t^M
\end{alignat*}
called \emph{methods}.
For brevity, we omit the rest of methods and refer to the Agda implementation for them.
\LT[noinline]{Add a link.}
We also have omitted the list of implicitly quantified variables: each variable $A$ implicitly quantified in a constructor $c$ has accordingly an $A$-indexed variable $A^M$ in $c^M$.

@@

Given an eliminator, the following clauses are derived to carry out the actual computation
\begin{alignat*}{5}
  \elim_{\Ctx}\;\emptyctx   & \reduce \emptyctx_{\Ctx}                                    && : \Ctx^M\;\emptyctx  \\
  \elim_{\Ctx}\;(\Gamma, A) & \reduce (\elim_{\Ctx}\,\Gamma),_{\Ctx}^{M} (\elim_{\Ty}\,A) && : \Ctx^M\;(\Gamma, A) \\
  \elim_{\Ty}\;\UU          & \reduce \UU^M && :\Ty^M\;\Gamma^M\;\UU \\
  \elim_{\Sub}\;\emptysub   & \reduce \emptysub^M_{\Sub} &&: \Sub^M\;\Gamma^M\;\emptyctx^M_{\Ctx}\;\emptyctx \\
                            & \vdots
\end{alignat*}
interpreting each constructor using the corresponding method inductively.

Interestingly, the key difference from its QII counterpart is that substitution operations $[\blank]_{\Ty}\blank$ and $[\blank]_{\Tm}$ are functions rather than constructors, meaning they cannot be eliminated inductively but instead become propositional identities.
%type substitution $[\blank]\blank$ was a constructor and $[\sigma]\,A$ was eliminated to $[\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A)$; similarly for term substitution.
\begin{proposition}\label{prop:provable-function-clauses}
  Given the motives and the methods, we have
\begin{alignat*}{5}
  \elim_{\Ty}[] & : \implicit{\dots} & \elim_{\Ty}\,([\sigma]_{\Ty}\,A) & = && [\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A) \\
  \elim_{\Tm}[] & : \implicit{\dots} & \elim_{\Tm}\,([\sigma]_{\Tm}\,t) & =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{\elim_{\Ty}[]}} && [\elim_{\Sub}\,\sigma]^M_{\Tm}\,(\elim_{\Tm}\,t)
\end{alignat*}
\end{proposition}
\LT{Add some remark regarding the proof?}
